---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start --> 
[![version](https://img.shields.io/badge/version-v0.1.0-blue)](https://github.com/ndpvh/predped/releases)
[![documentation](https://img.shields.io/badge/documentation-v0.1.0-blue)](https://ndpvh.github.io/predped)
[![build](https://github.com/ndpvh/predped/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ndpvh/predped/actions/workflows/R-CMD-check.yaml)
[![coverage](https://codecov.io/gh/ndpvh/predped/graph/badge.svg)](https://app.codecov.io/gh/ndpvh/predped)
[![license: GPL v3](https://img.shields.io/badge/license-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
<!-- badges: end -->

# predped: An R package to simulate the Minds for Mobile Agents (M4MA) pedestrian model

This project serves as a tool to simulate pedestrian movement using the Minds for Mobile Agents (M4MA) model. It contains several helper functions for people who wish to use the model to investigate pedestrian flow in their own settings. Before delving into how to use the package, we will first provide the reader with some background on the model.

## Background

Pedestrian models are popular tools to investigate how people navigate the complex world that we live in. While these models have certainly been useful to understand, for example, evacuation behavior, they most often assume that pedestrians are homogeneous -- that is, that all pedestrians are one and the same. Because of this assumption, these pedestrian models are not suited to capture the variety of walking behaviors we observe in most low-density situations, such as those observed in the supermarket, in the train station, or even on the street.

To alleviate these difficulties, Andrew Heathcote and Dora Matzke recently proposed M4MA to capture pedestrian movement. This model is based on the pedestrian model developed by Robin et al. (2009) and, similarly, proposes that pedestrian movement is determined on three levels: 

- _Strategic level_: Consists of planning where the pedestrian wants to go to and via which route. Typically, a pedestrian will have to fulfill a goal, which will serve as the destination of the route. To get to this goal, pedestrians will walk along so-called "path-points": Waypoints along which the pedestrian can walk, serving as intermediate steps to get to their destination. 
- _Tactical level_: Once their route is determined, pedestrians will still be able to adapt if their original plan does not seem feasible. For example, other pedestrians can block your route, which makes avoiding the crowd more attractive when trying to get to your goal. These adaptations are handled on this level.
- _Operational level_: This final level handles the low-level, moment-to-moment decisions about where to walk to. In practice, this constitutes a choice between one of 34 movement options, created by 11 angular directions in which a pedestrian can move (going from -72.5 to 72.5 degrees) and 3 changes in velocity (deceleration, constant speed, or acceleration), with the $34^{st}$ option representing stopping. The probability with which a pedestrian will move to a given location is determined by a utility function that combines several moving components, namely (a) preferred speed, (b) current direction, (c) goal direction, (d) interpersonal distance, and (e) blocked angles, as well as several social components, namely (a) following a person heading in the same direction (follow-the-leader) and (b) walking besides someone heading in the same direction (walk-besides-buddy).

Critically, and different from other pedestrian models, M4MA assigns each pedestrian a "personality", in that all pedestrians have their own unique values for the parameters that are defined on the operational level. These individual differences are implemented in two ways:

- _Qualitative differences_: We define several "archetypes" of pedestrians, each of which have different discrete values for their parameters. These archetypes were created through trial-and-error by the researchers in this project and are contained in _archetypes.csv_. In no way should these archetypes be interpreted as ground truth or real differences between individuals.
- _Quantitative differences_: Each of the parameter values that are defined in the archetypes CSV-file serves as a mean around which each pedestrian's actual parameter value fluctuates. Without going into detail, each archetype has their own matrix that contains standard deviations for each parameter on the diagonal and correlations between parameters on the off-diagonal. These matrices are stored in _archetypes\_sigma.Rds_.

## How to use

This package allows its users to simulate pedestrian behavior as expected by M4MA. Users can specify the environment, the characterstics the pedestrians, and the characteristics of the simulation itself (e.g., maximal number of agents, initial conditions,...), therefore allowing for a great variety of potential research questions that can be answered. For more information on the workflow of this package, see _Getting started_.

This package depends heavily on the [```m4ma```](https://github.com/m4ma/m4ma) package.

## Installation

One can install ```predped``` through the ```remotes``` package in the following way:

```{r}
#| eval = FALSE
remotes::install_github("ndpvh/predped")
```

To use the package, you should load it through ```library```:

```{r}
#| message = FALSE
library(predped)
```

## Getting started

Within ```predped```, we use a particular workflow. 

In the first step, you should define the environment in which you want pedestrians to walk around. For this, you will use the [`background`](https://ndpvh.github.io/reference/background-class.html) class. This S4 class contains slots for the ```shape``` (i.e., the shape of the room), the ```objects``` (i.e., which objects are contained within the room), and a potential ```entrance``` and/or ```exit```. The ```shape``` and ```objects``` should all consist of instances of the [`object`](https://ndpvh.github.io/reference/object-class.html) class, that is they should be either a [`rectangle`](https://ndpvh.github.io/reference/rectangle-class.html), [`polygon`](https://ndpvh.github.io/reference/polygon-class.html), or [`circle`](https://ndpvh.github.io/reference/circle-class.html). The ```entrance```/```exit``` should be either a numeric vector or a numeric matrix with two columns specifying their coordinates.

A simple example of an environment is the following circular room with a square object placed in the middle:

```{r}
setting <- background(
  shape = circle(
    center = c(0, 0), 
    radius = 2
  ), 
  objects = list(
    rectangle(
      center = c(0, 0), 
      size = c(1, 1)
    )
  ),
  entrance = c(-2, 0)
)
```

It is good practice to visualize what the environment looks like. For this, we can use the [`plot`](https://ndpvh.github.io/reference/plot.html) function:

```{r}
#| fig.width = 5, 
#| fig.height = 5.5,
#| fig.align = "center",
#| fig.alt = "One sees a plot visualizing a circular room with a square gray object in the middle. This represents the room in which the agents will walk around if the room is used for a simulation.",
#| out.width = "60%"
plot(setting)
```

Once an environment has been defined, on should link this environment with the characteristics of the agents who are expected to walk around in this environment. This is achieved through the definition of a [`predped`](https://ndpvh.github.io/reference/predped.html) model, another S4 class with slots ```setting``` (the environment we just created), ```parameters``` (dataframe of parameter values), ```archetypes``` (a character string of pedestrians to include), and ```weights``` (the probability of each archetype entering the room). 

Agent characteristics are defined through a `data.frame` that contains parameter values for a given "class" of people. One such dataframe is provided by us in _archetypes.csv_ and can be called in the package through the variable [`params_from_csv`](https://ndpvh.github.io/reference/params_from_csv.html) or through calling the function [`load_parameters()`](https://ndpvh.github.io/reference/load_parameters.html). In this example, we wish to use the `"BaselineEuropean"`, specifying the model as:

```{r}
model <- predped(
  id = "my model", 
  setting = setting, 
  archetypes = "BaselineEuropean"
)
```

Now that the model has been defined, we can now simulate pedestrian movement by calling the function [`simulate`](https://ndpvh.github.io/reference/simulate.html):

```{r}
#| cache = FALSE,
#| results = "hide"
set.seed(1)
trace <- simulate(
  model,
  max_agents = 25, 
  iterations = 50
)
```

The variable ```trace``` will be a list consisting of [`state`](https://ndpvh.github.io/reference/state.html)s of the environment, each state itself consisting of a copy of the environment (under slot ```setting```) and of a list of pedestrians walking around in the environment (under slot ```agents```). If you wish to visualize this trace, you can again use the [`plot`](https://ndpvh.github.io/reference/plot.html) function:

```{r}
#| cache = TRUE
plots <- plot(
  trace,
  print_progress = FALSE
)
```

The [`plot`](https://ndpvh.github.io/reference/plot.html) outputs a list of plots. For research purposes, it is useful to transform this list to a gif, which can be achieved by using the `gifski` package:

```{r}
#| eval = TRUE
gifski::save_gif(
  lapply(plots, print), 
  file.path("readme.gif"),
  delay = 1/10
)
```

Looking at the created `.gif` then gives us an idea of how the agents walked around in the room:

```{r}
#| eval = TRUE,
#| echo = FALSE,
#| out.width = "60%",
#| fig.align = "center",
#| fig.alt = "A .gif displaying how an agent comes in through the entrance on the left side, interacts with a goal on the left side of the object, completes it, and then moves to a next goal on the bottom side of the rectangle."
knitr::include_graphics(
  file.path("./readme.gif")
)
```

## Getting help

You can find the documentation for this package on its dedicated [documentation site](https://ndpvh.github.io/predped). This site includes additional information on the [theoretical background](https://ndpvh.github.io/predped/articles/theory.html), on [running minimal](https://ndpvh.github.io/predped/articles/simulation.html) and [advanced simulations](https://ndpvh.github.io/predped/articles/advanced_simulation.html), and on [estimating the model on data](https://ndpvh.github.io/predped/articles/estimation.html).

If you encounter a bug, you can report the bug with a minimal working example as an [Issue](https://github.com/ndpvh/predped/issues).

## Contribute

If you otherwise wish to contribute to this project, feel free to reach out to Niels Vanhasbroeck (niels.vanhasbroeck@gmail.com) and Andrew Heathcote (ajheathcote@gmail.com).

## Credits

The development of this package would not have been possible without the help of its many contributors. For the development of the ```m4ma``` package, we thank (in alphabetical order):

- [Andrew Heathcote](https://github.com/andrewheathcote)
- [Malte Lüken](https://github.com/maltelueken)
- [Charlotte Tanis](https://github.com/CharlotteTanis)

For the creation of the ```predped```, we thank several permanent project members (in alphabetical order):

- [Andrew Heathcote](https://github.com/andrewheathcote)
- [Niels Vanhasbroeck](https://github.com/ndpvh)

as well as many who worked with us temporarily (in alphabetical order):

- [Alexander Anderson](https://github.com/Alexanderson31)
- [Joris Goossen](https://github.com/JorisGoosen)
- [Malte Lüken](https://github.com/maltelueken)
- [Ece Yatıkçı](https://github.com/eceyatikci) 

## See also

For more information on the project, please see its dedicated section on the lab website: [https://www.ampl-psych.com/projects/minds-for-mobile-agents/](https://www.ampl-psych.com/projects/minds-for-mobile-agents/).

## References

Robin, T., Antonini, G., Beirlaire, M., & Cruz, J. (2009). Specification, estimation, and validation of a pedestrian walking behavior model. _Transportation Research Part B, 43_, 36-56. doi: [10.1016/j.trb.2008.06.010](doi.org/10.1016/j.trb.2008.06.010)