---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start --> 
[![version](https://img.shields.io/badge/version-v0.4.1-blue)](https://github.com/ndpvh/predped/releases)
[![documentation](https://img.shields.io/badge/documentation-v0.4.1-blue)](https://ndpvh.github.io/predped)
[![build](https://github.com/ndpvh/predped/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ndpvh/predped/actions/workflows/R-CMD-check.yaml)
[![coverage](https://codecov.io/gh/ndpvh/predped/graph/badge.svg)](https://app.codecov.io/gh/ndpvh/predped)
[![license: GPL v3](https://img.shields.io/badge/license-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
<!-- badges: end -->

# predped: An R package around the Minds for Mobile Agents (M4MA) pedestrian model

This project serves as a tool to simulate pedestrian movement using the Minds for Mobile Agents (M4MA) model. This package enables users to simulate from the model and estimate the parameters of the model on movement data, each tailored to their own setting of choice. 



## Background

Pedestrian models are popular tools to investigate how people navigate the complex world that we live in. Yet, most current models assume that pedestrians are all one and the same, making them solely applicable to study movement behavior in high-density situations like festivals or sporting events, but less so in low-density situations like supermarkets and train stations.

To capture movement behavior in both low- and high-density situations, Andrew Heathcote and Dora Matzke recently proposed the M4MA, proposing that pedestrian behavior is determined on three levels: 

- _Strategic level_: Consists of what the agent want to achieve and how they will navigate the space; 
- _Tactical level_: Consists of reacting to the environment and making tactical decisions regarding navigation in case of blockages;
- _Operational level_: Consists of the moment-to-moment step decisions the agent takes, encompassing changes in direction and/or speed at the lowest level.

Critically, M4MAs pedestrians have a particular "personality" reflected in unique parameters that guide the movement on the operational level. These individual differences are implemented in two ways:

- _Qualitative differences_: Each pedestrian belongs to a particular class of people, defining a particular pedestrian profile that one may encounter within the setting of interest;  
- _Quantitative differences_: Within each class, parameter values are subject to random variation, assigning each pedestrian with a unique combination of parameters.



## Installation

One can install `predped` via the `remotes` package:

```{r}
#| eval = FALSE
remotes::install_github("ndpvh/predped")
```

To use the package, one should load it through `library`:

```{r}
#| message = FALSE
library(predped)
```



## Getting started

To simulate data from the M4MA, `predped` requires the following workflow. 

First, one should define the environment in which the pedestrians will walk around, defined through the [`background`](https://ndpvh.github.io/reference/background-class.html) class. This S4 class defines the `shape` of the environment, , the `objects` that are contained in it, and a set of `entrance`s and/or `exit`s. The `shape` and `objects` are furthermore restricted to be instances of the [`object`](https://ndpvh.github.io/reference/object-class.html) class, that is they should be either a [`rectangle`](https://ndpvh.github.io/reference/rectangle-class.html), [`polygon`](https://ndpvh.github.io/reference/polygon-class.html), or [`circle`](https://ndpvh.github.io/reference/circle-class.html). 

A simple example of an environment is the following circular room with a square object placed in the middle:

```{r}
setting <- background(
  shape = circle(
    center = c(0, 0), 
    radius = 2
  ), 
  objects = list(
    rectangle(
      center = c(0, 0), 
      size = c(1, 1)
    )
  ),
  entrance = c(-2, 0)
)
```

It is good practice to visualize what the environment looks like. For this, we can use the [`plot`](https://ndpvh.github.io/reference/plot.html) function:

```{r plot}
#| fig.width = 5, 
#| fig.height = 5.5,
#| fig.align = "center",
#| fig.alt = "One sees a plot visualizing a circular room with a square gray object in the middle. This represents the room in which the agents will walk around if the room is used for a simulation.",
#| out.width = "60%"
plot(setting)
```

Once an environment has been defined, on should link this environment with the characteristics of the agents who are expected to walk around in this environment. This is achieved through defining an instance of the [`predped`](https://ndpvh.github.io/reference/predped.html) class, this time defining the `setting` or environment, the `archetypes` of the agents expected to walk in it, and the `weights` or probability with which one expects each archetype to be found within the space. 

Agent characteristics are defined through a `data.frame` that contains parameter values for a given "class" of people. One such dataframe is provided by us in _archetypes.csv_ and can be called in the package through the variable [`params_from_csv`](https://ndpvh.github.io/reference/params_from_csv.html) or through calling the function [`load_parameters()`](https://ndpvh.github.io/reference/load_parameters.html). In this example, we wish to use the `"BaselineEuropean"`, specifying the model as:

```{r}
model <- predped(
  id = "my model", 
  setting = setting, 
  archetypes = "BaselineEuropean"
)
```

Once the setting and `predped` model have been defined, one can simulate pedestrian movement as expected by the M4MA through calling the function [`simulate`](https://ndpvh.github.io/reference/simulate.html):

```{r}
#| cache = FALSE,
#| results = "hide"
set.seed(1)
trace <- simulate(
  model,
  max_agents = 25, 
  iterations = 50
)
```

The variable `trace` consists of a list of [`state`](https://ndpvh.github.io/reference/state.html)s of the environment, each state itself consisting of a copy of the environment (under slot `setting`) and of a list of pedestrians walking around in the environment (under slot `agents`). To visualize this trace, one can again use the [`plot`](https://ndpvh.github.io/reference/plot.html) function:

```{r}
#| cache = TRUE
plots <- plot(
  trace,
  print_progress = FALSE
)
```

The [`plot`](https://ndpvh.github.io/reference/plot.html) outputs a list of plots. For research purposes, it is useful to transform this list to a gif, which can be achieved by using the `gifski` package:

```{r}
#| eval = FALSE
gifski::save_gif(
  lapply(plots, print), 
  file.path("readme.gif"),
  delay = 1/10
)
```

```{r}
#| eval = TRUE, 
#| echo = FALSE 
gifski::save_gif(
  lapply(plots, print), 
  file.path("man/figures/readme.gif"),
  delay = 1/10
)
```

Looking at the created `.gif` then gives us an idea of how the agents walked around in the room:

```{r}
#| eval = TRUE,
#| echo = FALSE,
#| out.width = "60%",
#| fig.align = "center",
#| fig.alt = "A .gif displaying how an agent comes in through the entrance on the left side, interacts with a goal on the left side of the object, completes it, and then moves to a next goal on the bottom side of the rectangle."
knitr::include_graphics(
  file.path("man/figures/readme.gif")
)
```



## Getting help

You can find the documentation for this package on its dedicated [documentation site](https://ndpvh.github.io/predped). For an overview of the logic within the package, we refer the interested reader to the [Getting Started](https://ndpvh.github.io/predped/articles/getting_started.html) page on this site. On this site, we furthermore include information on the [theoretical background](https://ndpvh.github.io/predped/articles/theory.html), on [running minimal](https://ndpvh.github.io/predped/articles/simulation.html) and [advanced simulations](https://ndpvh.github.io/predped/articles/advanced_simulation.html), and on [estimating the model on data](https://ndpvh.github.io/predped/articles/estimation.html).

If you encounter a bug, you can report the bug with a minimal working example as an [Issue](https://github.com/ndpvh/predped/issues).



## Contribute

If you otherwise wish to contribute to this project, feel free to reach out to Niels Vanhasbroeck (niels.vanhasbroeck@gmail.com) and Andrew Heathcote (ajheathcote@gmail.com).



## Credits

The development of this package would not have been possible without the help of its many contributors. For the development of the `m4ma` package, we thank (in alphabetical order):

- [Andrew Heathcote](https://github.com/andrewheathcote)
- [Malte Lüken](https://github.com/maltelueken)
- [Charlotte Tanis](https://github.com/CharlotteTanis)

For the creation of the ```predped```, we thank its permanent project members (in alphabetical order):

- [Andrew Heathcote](https://github.com/andrewheathcote)
- [Niels Vanhasbroeck](https://github.com/ndpvh)

as well as those who worked with us temporarily (in alphabetical order):

- [Alexander Anderson](https://github.com/Alexanderson31)
- [Joris Goossen](https://github.com/JorisGoosen)
- [Jakub Jurak](https://github.com/kubojurak)
- [Malte Lüken](https://github.com/maltelueken)
- [Ece Yatıkçı](https://github.com/eceyatikci) 



## See also

For more information on the project, please see its dedicated section on the lab website: [https://www.ampl-psych.com/projects/minds-for-mobile-agents/](https://www.ampl-psych.com/projects/minds-for-mobile-agents/).
