---
title: "Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#| echo = FALSE,
#| message = FALSE
library(predped)
library(nloptr)
```

One of the major advances of the Minds for Mobile Agents model (M4MA) is the ability to estimate its operational parameters on movement data of sufficient quality. In this vignette, we will illustrate this through two examples, one starting from the `trace` output from the [`simulate`](https://ndpvh.github.io/predped/reference/simulate) function and the other starting from a `data.frame` containing measured movement data. 

Note that for the estimation, `predped` only concerns itself with computing the utilities and associated probabilities of each expected movement decision. This means that we need to rely on external packages to perform the estimation procedure, using an optimization procedure to maximize the likelihood of the data under a set of parameters. Within this vignette, we use the `nloptr` package to perform this optimization procedure (Johnson, 2008). To install and use this package, one can execute the following two commands:

```{r}
#| eval = FALSE
install.packages("nloptr")
library(nloptr)
```

For more information on the package and the nlopt library, we refer the interested reader to the dedicated documentation sites (`nloptr`: https://astamm.github.io/nloptr; NLopt: https://nlopt.readthedocs.io/en/latest).



# Disclaimer

Note that data quality is determined by factors that commonly affect estimation in most domains, including: 1) the level of noise added by measurement (less is better!), 2) the lengths movements observed (longer is better!), and 3) the type of interactions among agents observed (some utilities are only identified if the behaviors they govern are observed in sufficient quantity). Experimentation is required to determine what level of quality is adequate in a specific scenario. Also note that the computation of the likelihood can become slow in large scenarios, so estimation may be practically difficult without large computing resources.


# Starting from a Trace 

Consider a trace that has been saved in the variable `trace`, containing a simulation in which people walk around in a minimal supermarket with some additional flow indicators to prevent blockages (see [_Advanced Simulations_](https://ndpvh.github.io/predped/articles/advanced_simulation.html)):

```{r}
#| echo = FALSE,
#| eval = FALSE
# Create a simplified version of a supermarket
supermarket <- background(
  # Shaking things up: Create a polygon environment
  shape = polygon(
    points = cbind(
      c(0, 0, 5, 5, 10, 10), 
      c(0, 5, 5, 10, 10, 0)
    )
  ), 

  # Objects in the environment, consisting of 
  # some shelves and a gated entrance. Gates are
  # again not interactable.
  objects = list(
    # Top shelves
    rectangle(center = c(7.5, 7), size = c(0.7, 4)),
    rectangle(center = c(6, 7), size = c(0.7, 4)),
    rectangle(center = c(9, 7), size = c(0.7, 4)),

    # Bottom shelves
    rectangle(center = c(3, 2.5), size = c(4, 0.7)),
    rectangle(center = c(3, 1), size = c(4, 0.7)),

    # Central box
    rectangle(center = c(7.5, 2.5), size = c(2, 2)),

    # Cash register and gate
    polygon(
      points = cbind(
        c(1, 2.25, 2.25, 2.5, 2.5, 1), 
        c(3.75, 3.75, 4.25, 4.25, 3.5, 3.5)
      ), 
      interactable = FALSE
    ),
    rectangle(
      center = c(3.5, 4.25), 
      size = c(0.25, 1.5), 
      interactable = FALSE
    )
  ),

  # Entrance and exit are at the cash register
  entrance = c(1, 5),

  # Limited access so that agents have to pass the cash 
  # register beforehand
  limited_access = list(
    # Entrances
    segment(from = c(1, 3.5), to = c(0, 3.5)),
    segment(from = c(2.5, 4.25), to = c(3.375, 4.25)),

    # Bottom shelves
    segment(from = c(5, 2.5), to = c(5, 1)),
    segment(from = c(1, 2.5), to = c(1, 3.5)),
    segment(from = c(1, 0), to = c(1, 0.65)),
    segment(from = c(1, 2.5), to = c(1, 1)),
    segment(from = c(5, 2.5), to = c(5, 3.5)),
    segment(from = c(5, 0), to = c(5, 0.65)),

    # Top shelves
    segment(from = c(5, 9), to = c(5.65, 9)),
    segment(from = c(9.3, 9), to = c(10, 9)),
    segment(from = c(7.10, 5), to = c(6.35, 5)),
    segment(from = c(8.7, 5), to = c(7.75, 5)),
    segment(from = c(5, 5), to = c(5.65, 5)),
    segment(from = c(9.3, 5), to = c(10, 5)),
    segment(from = c(7.10, 9), to = c(6.35, 9)),
    segment(from = c(8.7, 9), to = c(7.75, 9))
  )
)

# Plot the supermarket
plot(
  supermarket, 
  segment.hjust = 1
)

# Set seed
set.seed(1)

# Connect environment to agent characteristics
my_predped <- predped(
  setting = supermarket, 
  archetypes = c("BaselineEuropean", "BigRushingDutch")
)

# Run the simulation and provide our function 
# to the `fx` argument
trace <- simulate(
  my_predped, 
  max_agents = 5, 
  add_agent = 20, 
  iterations = 200,
  print_iteration = TRUE,
  individual_differences = TRUE
)

# Create plots and GIF
plt <- plot(
  trace, 
  plot_segment = FALSE
)

# Save to a gif
gifski::save_gif(
  lapply(plt[1:200], print),
  file.path("man", "figures", "trace.gif"),
  delay = 1/10
)

# Save the trace itself
saveRDS(
  trace, 
  file.path("man", "data", "trace.Rds")
)
```

```{r}
#| echo = FALSE,
#| fig.align = "center",
#| fig.alt = "Visualization of the minimal supermarket and the agents that walk around in it. The first 200 iterations of movements of the agents within this supermarket are visualized in this GIF."
knitr::include_graphics(
  file.path("..", "man", "figures", "trace.gif")
)

trace <- readRDS(
  file.path("..", "man", "data", "trace.Rds")
)

```

Then we need to perform the following steps to estimate the parameters of the M4MA to the data contained within this `trace`.

First, one will have to transform the `trace` to a `data.frame`, which can be achieved with the [`unpack_trace`](https://ndpvh.github.io/predped/reference/unpack_trace.html) function:

```{r}
data <- unpack_trace(trace)
```

The variable `data` contains information in a variety of columns, namely:

```{r}
colnames(data)
```

Of these, some indicate attributes of the agent at a particular time (e.g., position in `x` and `y`, speed and orientation in `speed` and `orientation`) but also the cell that they selected (`cell`) and variables that are relevant for computing the utility of that decision (e.g., `check`, `ps_speed`, `gd_angle`,...). The contents of the latter set of columns depend heavily on the requirements of the utility functions defined in the [`m4ma`](https://github.com/m4ma/m4ma) package and we refer the interested reader to its documentation for a full explanation (https://research-software-directory.org/software/m4ma).

Second, we need to delete data points that are not relevant for estimation. Given that the M4MAs operational parameters guide movement, we can only use those data points at which agents were moving around (except in cases where stopping is determined by the `stop_utility` parameter). Within a `trace`, the easiest way in which to spot whether this is the case is by filtering out any rows in which the utility variables are `NA`, indicating that they were not used during that iteration:

```{r}
dim(data)

data <- data[!is.na(data$check), ]

dim(data)
```

Note that this step is not necessary for the code to run, but that a warning would be thrown in case `NA` are left in the data. 

Finally, once the data has been filtered to contain only utility-governed data, we can compute the likelihood of that data under a particular set of operational parameters. Calling this likelihood $L$, then one can compute the negative log-likelihood $-\log(L)$ (i.e., a value for which smaller magnitudes indicate a better fit) with the [`mll`](https://ndpvh.github.io/predped/reference/mll.html) function as follows:

```{r}
# Define a set of parameters to be evaluated. For illustration purposes, we use
# the parameters of the DrunkAussie for this purpose. Note that we delete the 
# agent's type name and their color from the parameters. Furthermore note that 
# we need to transform the parameters to a numeric vector.
params <- load_parameters()$params_archetypes
params <- params[params$name == "DrunkAussie", -c(1, 2)] |>
  unlist() |>
  as.numeric()

# Compute the likelihood of the data under this parameter set
mll(
  data, 
  params,
  transform = FALSE,
  summed = TRUE
)
```

The output of this call to [`mll`](https://ndpvh.github.io/predped/reference/mll.html) is a named vector displaying the summed negative log-likelihood of the data under the provided parameters. 

In this piece of code, we specified the argument `transform = FALSE` to communicate that the provided parameters are on the bounded scale rather than the unbounded scale (see _Parameters_ in the vignette on [_Simulations_](https://ndpvh.github.io/predped/articles/simulation.html)). When using an optimizer, it is usually better to let it work on an unbounded scale (i.e., the full real line), in which case you can specify `transform = TRUE`. Additionally, we specified `summed = TRUE` to get the summed value for the negative log-likelihood across all data points of a particular agent. If you wish to see the values of the negative log-likelihood for each data point individually, you can set `summed = FALSE`.

The call to the [`mll`](https://ndpvh.github.io/predped/reference/mll.html) function will serve as the basic building block of the estimation procedure, which we dive more into in the next section.



## Bounded Optimization 

Typically, we want to estimate a set of unknown parameters best describing our data. To achieve this via the `nloptr` package, we need to first define an _objective function_, that is a function that outputs a value to be minimized by the optimizer through changing a vector of parameters that serve as an input to this function. In our case, the objective function will be a wrapper around the [`mll`](https://ndpvh.github.io/predped/reference/mll.html) function, summing negative log-likelihoods across agents into one single `objective` value:

```{r}
objective_function <- function(parameters) {
  return(
    sum(
      mll(
        data, 
        parameters,
        transform = FALSE,
        summed = TRUE
      )
    )
  )
}
```

Note that this function only takes in a single argument `parameters`: The content of the `data` that is included in the call to `mll` is assumed to be a global variable. Furthermore note that for our purposes, we assume that the provided parameters fall on the bounded scale and therefore don't need to be transformed.

Once the objective function is defined, we can call the `nloptr` function of the `nloptr` package to perform the estimation procedure. Using the dividing rectangles algorithm (Jones et al., 1993), for example, we get: 

```{r}
#| cache = TRUE
# Extract the bounds of the parameters
bounds <- load_parameters()$params_bounds

# Define the initial condition for the algorithm. Take the middle of the 
# interval defined by the bounds
x0 <- rowMeans(bounds)

# Perform the estimation procedure
fit <- nloptr(
  x0,
  objective_function, 
  lb = bounds[, 1],
  ub = bounds[, 2],
  opts = list(
    algorithm = "NLOPT_GN_DIRECT",
    maxeval = 1000,
    xtol_abs = 1e-4,
    ftol_abs = 1e-4
  )
)

# Print the resulting fit
fit
```



## Individual Parameters 

Up to now, we have assumed that each of the participants has the same parameter set underlying their movement data. More realistically, though, individuals differ in their walking tendencies, requiring us to estimate different parameters for each individual within our data. 

To achieve this, we have to provide a matrix of parameter values to the [`mll`](https://ndpvh.github.io/predped/reference/mll.html) function with rows indicating the individuals and the columns the parameters of interest. In practice, we therefore need to change the objective function so that:

```{r}
# Get the number of participants in the study
n <- length(unique(data$id))

# Define the objective function
objective_function <- function(parameters) {
  # Transform the parameters to the required format
  parameters <- matrix(
    parameters,
    nrow = n, 
    ncol = 33
  )

  # Compute the objective
  return(
    sum(
      mll(
        data, 
        parameters,
        transform = FALSE,
        summed = TRUE
      )
    )
  )
}
```

Then, it's just a matter of changing the initial conditions `x0` and the bounds `lb` and `ub` in the `nloptr` function so that it matches the objective function, in this case:

```{r}
#| cache = TRUE
# Extract the bounds of the parameters
bounds <- load_parameters()$params_bounds |>
  rep(each = n) |>
  matrix(nrow = 33 * n, ncol = 2)

# Define the initial condition for the algorithm. Take the middle of the 
# interval defined by the bounds
x0 <- rowMeans(bounds)

# Perform the estimation procedure
fit <- nloptr(
  x0,
  objective_function, 
  lb = bounds[, 1],
  ub = bounds[, 2],
  opts = list(
    algorithm = "NLOPT_GN_DIRECT",
    maxeval = 1000,
    xtol_abs = 1e-4,
    ftol_abs = 1e-4
  )
)

# Print the resulting fit
fit
```

In this case, it may be useful to transform the parameters back to the matrix introduced in the objective function and provide it with matching column names denoting the parameters and row names denoting the participants themselves:

```{r}
# Get the parameter names: By default the same order as the one in bounds
params <- load_parameters()$params_bounds |>
  rownames()

# Get the id names: The same order as the ones output by mll
ids <- mll(data, bounds[, 1], transform = FALSE) |>
  names()

# Transform the result to a matrix with the provided column and row names
fit$solution |>
  matrix(nrow = length(ids), ncol = length(params)) |>
  `rownames<-` (ids) |>
  `colnames<-` (params)
```



## Unbounded Estimation

The algorithm that we used before requires us to impose bounds on the parameters. Often, however, optimization works best if the entire real line can be searched. In this case, we change the argument `transform` to `TRUE` and change the call to `nloptr` as follows:

```{r}
#| cache = TRUE
# Change the objective function to operate on the real scale
objective_function <- function(parameters) {
  return(
    sum(
      mll(
        data, 
        parameters,
        transform = TRUE,
        summed = TRUE
      )
    )
  )
}

# Change the initial condition
x0 <- numeric(33)

# Perform the estimation procedure with an algorithm that does not require 
# bounds to be specified
fit <- nloptr(
  x0,
  objective_function, 
  opts = list(
    algorithm = "NLOPT_LN_SBPLX",
    maxeval = 1000,
    xtol_abs = 1e-4,
    ftol_abs = 1e-4
  )
)

# Print the resulting fit
fit
```

Importantly, the parameters now have been estimated to be on the real line, yet the parameters of the M4MA need are naturally defined on  bounded scales. To transform the parameters to their natural values, we use the [`to_bounded`](https://ndpvh.github.io/predped/reference/to_bounded.html) function:

```{r}
# Get the bounds of the parameters
bounds <- load_parameters()$params_bounds

# Extract the parameters and transform to a data.frame. Expected by the 
# to_bounded function
params <- fit$solution |>
  matrix(nrow = 1) |>
  as.data.frame() |>
  `colnames<-` (rownames(bounds))

# Use to_bounded to transform the parameters
transformed <- to_bounded(
  params, 
  bounds
)
transformed
```



# Starting from Data

Within `predped`, we also allow users to estimate the M4MA on their data directly. Consider the following data set taken from a simplified train station, for example (see [_Advanced Simulations_](https://ndpvh.github.io/predped/articles/advnaced_simulation.html)):

```{r}
#| echo = FALSE,
#| include = FALSE
train_station <- background(
  # Define a space of size (10, 5) meters
  shape = rectangle(
    center = c(0, 0), 
    size = c(10, 5)
  ), 
  
  # Define some gates at the entrances. These gates are 
  # not interactable.
  objects = list(
    # Gates on the left
    rectangle(
      center = c(-3, 0), 
      size = c(0.6, 1.3),
      interactable = FALSE
    ), 
    rectangle(
      center = c(-3, 1.9), 
      size = c(0.6, 1.2),
      interactable = FALSE
    ),
    rectangle(
      center = c(-3, -1.9), 
      size = c(0.6, 1.2),
      interactable = FALSE
    ),

    # Gates on the right
    rectangle(
      center = c(3, 0), 
      size = c(0.6, 1.3),
      interactable = FALSE
    ), 
    rectangle(
      center = c(3, 1.9), 
      size = c(0.6, 1.2),
      interactable = FALSE
    ),
    rectangle(
      center = c(3, -1.9), 
      size = c(0.6, 1.2),
      interactable = FALSE
    )
  ),

  # Define the entrances themselves. The two entrances
  # at the top and bottom of the space represent the 
  # stairs agents can use to exit the train tracks. 
  # The two entrances at the left and right represent
  # the entrances of the train station.
  entrance = rbind(
      # Top and bottom
      c(0, 2.5), 
      c(0, -2.5),

      # Left and right
      c(-5, 0),
      c(5, 0)
  )
)

# Plot the train station
plot(
  train_station, 
  segment.hjust = 1
)
```

```{r}
#| echo = FALSE,
#| eval = FALSE
# Set seed
set.seed(1)

# Connect environment to agent characteristics
my_predped <- predped(
  setting = train_station, 
  archetypes = "BaselineEuropean"
)

# Run the simulation and provide our function 
# to the `fx` argument
trace <- simulate(
  my_predped, 
  max_agents = 10, 
  add_agent = 5, 
  iterations = 200,
  print_iteration = TRUE,
  individual_differences = TRUE
)

# Create plots and GIF
plt <- plot(
  trace, 
  plot_segment = FALSE
)

# Save to a gif
gifski::save_gif(
  lapply(plt[1:200], print),
  file.path("man", "figures", "trace_train.gif"),
  delay = 1/10
)

# Save the trace itself
saveRDS(
  trace, 
  file.path("man", "data", "trace_train.Rds")
)

# Create a dataset out of this
data <- time_series(trace)[, c("time", "id", "x", "y")]

data$id <- factor(data$id) |>
  as.numeric()
data$time <- jitter(data$time)
colnames(data) <- c("time", "passenger", "x", "y")

write.table(
  data, 
  file.path("man", "data", "data.txt"),
  sep = ","
)
```

```{r}
#| echo = FALSE
data <- read.table(
  file.path("..", "man", "data", "data.txt"),
  sep = ","
)
```

```{r}
head(data)
```

In this data set, we measured the positions (`x` and `y`) of different passengers (`passenger`) at several time points (`time`). To go from this data set to one that we can use for the estimation procedure, we need to take the following steps.

First, we need the data set to conform to the standards of `predped`. This means that:

- The `time` variable needs to show discrete jumps conforming the `time_step` defined by the M4MA (by default, half a second);
- A variable `iteration` needs to be added showing the iteration at which the positions were measured;
- The `passenger` variable needs to be renamed to the `predped` internal `id`.

Addressing the first change, this amounts to binning the data within a particular time window, aggregating the positions within that window through, for example, taking a mean, as we do in the next bit of code:

```{r}
# Define the bins, assuming the time variable is specified in seconds
bins <- seq(
  min(data$time), 
  max(data$time) + 0.5, 
  by = 0.5
)

# Loop over the bins and aggregate the data within those bins. Take the average
# of the bin as the time indication. Do this for each person separately.
data <- lapply(
  unique(data$passenger),
  function(x) {
    data <- lapply(
      1:(length(bins) - 1),
      function(i) {
        # Select the relevant data, it being the data where the time lies within
        # a particular bin and the passenger is x
        idx <- data$time <= bins[i + 1] & data$time >= bins[i] & data$passenger == x
        selection <- data[idx, ]

        # Aggregate and return a single row of data
        if(nrow(selection) > 0) {
          return(
            data.frame(
              time = (bins[i] + bins[i + 1]) / 2,
              passenger = x,
              x = mean(selection$x),
              y = mean(selection$y)
            )
          )
        } else {
          return(NULL)
        }
      }
    )

    # Bind together and return
    return(do.call("rbind", data))
  }
)

# Bind all rows within the list of binned data together
data <- do.call("rbind", data)

# Order based on time
data <- data[order(data$time), ]

head(data)
```

The `iteration` variable requires integer values denoting the bin in which the position was measured. To compute this variable, we can perform the following computation:

```{r}
data$iteration <- (data$time - min(data$time)) / 0.5 + 1
head(data)
```

The column `passenger` can be renamed to `id` in the following way:

```{r}
colnames(data)[2] <- "id"
head(data)
```

Now that the dataset conforms to `predped`s standards, we have to compute the values of the input variables for the input functions. For this, we first need to translate the environment in which the data were collected to an instance of the [`background`](https://ndpvh.github.io/predped/reference/background.html) to the best of our abilities. In this case, we collected the data in a simplified train station that looks as follows:

```{r}
#| echo = FALSE,
#| fig.align = "center",
#| fig.alt = "Visualization of the simplified train station in which the data were collected. Gates can be found on the left and right side of the space, and four exits and entrances can be found in the middle of each side of the rectangular space."
plot(
  train_station,
  segment.hjust = 1
)
```

Note that when there are restrictions in flow, this is ideally included in the creation of this background (see [_Advanced Simulations_](https://ndpvh.github.io/predped/reference/advanced_simulation.html)). 

Additionally, we need to include information on the goals that the people were trying to achieve within this environment, information that is necessary due to the M4MAs assumption that movement within space is guided by goals. Specifically, we need to define the position of the goal as well as attach an identifier to this goal, each respectively contained in the `goal_x`, `goal_y`, and `goal_id` columns in the data. In this case, we know that each person was walking towards a particular exit of the space. Based on this information, we add the required columns as follows:

```{r}
# Perform the mapping of a person with their goal position
mapping <- list(
  "exit 1" = c(1, 3, 12, 14, 21, 23, 28, 32, 33), 
  "exit 2" = c(2, 4, 5, 6, 9, 11, 16, 25),
  "exit 3" = c(7, 13, 18, 22, 24, 26, 30),
  "exit 4" = c(8, 10, 15, 17, 19, 20, 27, 29, 21)
)
coordinates <- list(
  "exit 1" = c(0, -2.5),
  "exit 2" = c(-5, 0),
  "exit 3" = c(0, 2.5),
  "exit 4" = c(5, 0)
)

# Add placeholders in the dataset for the columns to be filled later
data$goal_x <- 0
data$goal_y <- 0
data$goal_id <- " "

# Loop over the different mappings and fill the columns of interest
for(i in seq_along(mapping)) {
  # Define the rows to be filled based on the passengers moving to a particular
  # location
  idx <- data$id %in% mapping[[i]]

  # Fill the columns based on the exit's information
  exit <- names(mapping)[i]

  data$goal_x[idx] <- coordinates[[exit]][1]
  data$goal_y[idx] <- coordinates[[exit]][2]
  data$goal_id[idx] <- paste("moving to", exit)
}

head(data)
```

Once the setting and the people's goals have been defined, we can compute the input variables for the utility functions through the [`compute_utility_variables`](https://ndpvh.github.io/predped/reference/compute_utility_variables.html) function as follows:

```{r}
#| cache = TRUE
utility_data <- compute_utility_variables(
  data,
  train_station
)
colnames(utility_data)
```

As shown, the result contains all of the same columns as the ones found in the transformed `trace` in the previous section. From here on, the steps to take in the estimation routine are thus the same and won't be repeated.




# References 

Johnson, S. G. (2008). _The NLopt nonlinear-optimization package._ https://github.com/stevengj/nlopt

Jones, D. R., Perttunen, C. D., & Stuckmann, B. E. (1993). Lipschitzian optimization without the lipschitz constant. _Journal of Optimization Theory and Applications, 79_, 157.
