---
title: "Simulations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#| echo = FALSE,
#| message = FALSE
library(predped)
```

One of the major functionalities of the `predped` package is allowing users to simulate data according to the Minds for Mobile Agents pedestrian model (M4MA). In this section, we will detail the workflow for carrying out basic simulations. Please see the vignette on [_Advanced Simulations_](https://ndpvh.github.io/predped/articles/advanced_simulation.html) for ways in which you can personalize the simulations in `predped`.



# Defining an Environment 

First, one should define the environment in which the agents are expected to walk around. Within `predped`, we currently allow for the creation of a space at a single floor, which may represent a single room (e.g., an office or a supermarket) or a collection of connected rooms (e.g., a tunnel under a train station containing several stores). To define such a space, one should create an instance of the [`background`](https://ndpvh.github.io/reference/background.html) class, defining the following attributes: 

- `shape`: An instance of the [`object`](https://ndpvh.github.io/reference/object.html) class that defines the shape of the setting;
- `objects`: A list of [`object`](https://ndpvh.github.io/reference/object.html)s defining the obstacles in the setting;
- `entrance`, `exit`: Numeric matrices defining the entrances and/or exits of the space;
- `limited_access`: A list of [`segment`](https://ndpvh.github.io/reference/segment.html)s that restrict the movement of the agents within the space (see [_Advanced Simulations_](https://ndpvh.github.io/predped/articles/advanced_simulation.html)).

As an example, let's define a rectangular environment that represents an office:

```{r}
# Recreate our office
my_background <- background(
  # Shape of the environment
  shape = rectangle(
    center = c(0, 0), 
    size = c(4.5, 5)
  ), 

  # Objects contained within the environment
  objects = list(
    # Desks
    rectangle(
      center = c(-0.85, 0), 
      size = c(2.4, 1.6)
    ), 
    # Cabinets
    rectangle(
      center = c(-1, -2.3), 
      size = c(1.2, 0.4)
    ), 
    rectangle(
      center = c(2.05, -1.9), 
      size = c(0.4, 1.2)
    ), 
    # Big bookcase
    rectangle(
      center = c(0.35, 2.275), 
      size = c(3, 0.45)
    ),
    # Plants
    circle(
      center = c(-1.75, 2.275), 
      radius = 0.2
    ),
    circle(
      center = c(-2, -2.3), 
      radius = 0.15
    )
  ), 

  # Location of the entrance to our office
  entrance = c(2.25, 1.3)
)                                               
```

To visualize what this setting looks like, we can use the [`plot`](https://ndpvh.github.io/predped/reference/plot.html) method:

```{r}
#| fig.align = "center",
#| out.width = "100%",
#| fig.alt = "One sees the visualization our office, having a rectangular desk on the middle-left, some rectangular bookcases at the bottom-left, bottom-right, and top, and some round plants on the left of the bottom-left and top bookcases. All objects are filled with a gray color, while the space on which agents can walk is white. The location of the entrance is visualized through a half-circle on the top-right of the space."
plot(my_background)
```

It is often useful to create your setting by going back and forth between adding an object and plotting the result, especially when the environment contains many different objects.



## Interactability

By default, `predped` assumes that all objects in the environment can be interacted with. However, as this assumption may not apply to all situations, we also allow users to limit the interactability of the objects in their environment. This is done in two ways. 

First, users can make an object completely non-interactable by setting the `interactable` argument of the constructors for `rectangle`s, `polygon`s, and `circle`s to `FALSE`, such as in the following examples:

```{r}
# Turn off interactability of the objects
my_rectangle <- rectangle(
  center = c(0, 0), 
  size = c(1, 1), 
  interactable = FALSE
)

my_polygon <- polygon(
  points = rbind(
    c(1, 1), 
    c(1, -1), 
    c(-1, -1), 
    c(-1, 1)
  ),
  interactable = FALSE
)

my_circle <- circle(
  center = c(0, 0), 
  radius = 1, 
  interactable = FALSE
)
```

Please note that if none of the objects in a given `background` are interactable, agents will enter the space and immediately leave again through one of the provided exits. 

Another way in which interactability can be limited is through specifying the `forbidden_edges` argument in the same constructors. Essentially, this argument tells `predped` which regions of the objects cannot be interacted with. For `polygon`s and `rectangle`s, this amounts to providing the indices of the edges that cannot be interacted with, for example:

```{r}
my_polygon <- polygon(
  points = rbind(
    c(1, 1), 
    c(1, -1), 
    c(-1, -1), 
    c(-1, 1)
  ), 
  forbidden = c(1, 3)
)
```

and defining $P_1 = (1, 1)$, $P_2 = (1, -1)$, $P_3 = (-1, -1)$, and $P_4 = (-1, 1)$, then the edges that cannot be interacted with are the lines $\overset{\leftrightarrow}{P_1 P_2}$ and $\overset{\leftrightarrow}{P_3 P_4}$. This can be seen by executing the following code:

```{r}
# Create the edges by extracting the coordinates that make up the polygon and 
# then combining them in fixed order
coords <- points(my_polygon)
edges <- cbind(coords, coords[c(2:nrow(coords), 1), ])

# Identify those edges that cannot contain a goal
idx <- forbidden(my_polygon)
edges[idx, ]
```

The same ideas apply to the `rectangle`, using the output of the [`points`](https://ndpvh.github.io/predped/reference/points.html) method to create the relevant edges.

For `circle`s, there are no discrete edges that can be indexed like for `polygon`s and `rectangle`s. Within `predped`, we therefore use angles to specify the locations on the `circle` that cannot be interacted with, taking the shape of a matrix containing the starting and the ending angles of the interval in which no goal can be contained. Consider the following example, where we create a circle that cannot contain a goal in the region $[0, \pi]$.

```{r}
# Define a circle with forbidden angles ranging from 
# 0 to pi
my_circle <- circle(
  center = c(0, 0), 
  radius = 1, 
  forbidden = matrix(c(0, pi), nrow = 1)
)
```

Note that the angles should be specified in radians, not in degrees.

To simplify the specification of the `forbidden` argument, users can visualize the non-interactable locations when visualizing the environment. Redefining our office to include non-interactable locations:

```{r}
# Adjust the office to contain forbidden edges
my_background <- background(
  shape = rectangle(
    center = c(0, 0), 
    size = c(4.5, 5)
  ), 

  objects = list(
    rectangle(
      center = c(-0.85, 0), 
      size = c(2.4, 1.6),
      forbidden = 1
    ), 
    rectangle(
      center = c(-1, -2.3), 
      size = c(1.2, 0.4), 
      forbidden = c(1, 3, 4)
    ), 
    rectangle(
      center = c(2.05, -1.9), 
      size = c(0.4, 1.2), 
      forbidden = 2:4
    ), 
    rectangle(
      center = c(0.35, 2.275), 
      size = c(3, 0.45), 
      forbidden = 1:3
    ),
    circle(
      center = c(-1.75, 2.275), 
      radius = 0.2, 
      forbidden = rbind(
        c(0, 5 * pi / 4),
        c(7 * pi / 4, 2 * pi)
      )
    ),
    circle(
      center = c(-2, -2.3), 
      radius = 0.15,
      forbidden = rbind(
        c(0, pi / 4),
        c(3 * pi / 4, 2 * pi)
      )
    )
  ), 

  entrance = c(2.25, 1.3)
)
```

We can then visualize these locations by calling the `plot` method:

```{r}
#| fig.align = "center",
#| out.width = "100%",
#| fig.alt = "One sees the visualization our office, having a rectangular desk on the middle-left, some rectangular bookcases at the bottom-left, bottom-right, and top, and some round plants on the left of the bottom-left and top bookcases. All objects are filled with a gray color, while the space on which agents can walk is white. The location of the entrance is visualized through a half-circle on the top-right of the space."
# Plot the background with forbidden locations
plot(
  my_background, 
  plot_forbidden = TRUE,
  forbidden.color = "red"
)
```

In this plot, non-interactable locations are visualized in red. 

Note that interactability of an object will automatically be limited by where the objects are placed in the environment. More concretely, agents cannot interact with locations that they cannot reach, such as sides of an object that stand against another object or a wall. You do not need to specify this explicitly, though it may help during the setup of your simulation study.



## Debugging

There are some known issues or limitations regarding the [`background`](https://ndpvh.github.io/predped/reference/background.html) class that users should be aware off when creating their own environments. These are included here:

- Instances of the [`segment`](https://ndpvh.github.io/predped/reference/segment.html) class cannot be included as an object in the `objects` argument, as they solely exist to restrict the directionality of the pedestrian flow within an environment.



# Defining Agents

Once an environment has been defined, one should define the characteristics of the agents that are expected to walk around in this environment. An agent is an instance of the [`predped`](https://ndpvh.github.io/predped/reference/predped.html) class. It has the following attributes:

- `setting`: An instance of the [`background`](https://ndpvh.github.io/predped/reference/predped.html) class defining the environment in which agents should walk around;
-  `archetypes`: A character vector containing the names of the types of agents you want to include;
- `weights`: A numerical vector denoting the probability with which each agent type can be found within the environment;
- `parameters`: A list containing all necessary parameter information for `predped` to be able to simulate distinct agents.    

One can create an instance of the `predped` class through its constructor, for example by calling:

```{r}
# Create instance of predped class with our office and the default 
# parameters
my_predped <- predped(setting = my_background)    
my_predped
```

This specification connects the previously defined environment to the type of agents that should walk around in it. By default, all archetypes (i.e., types of agents) defined in the project's \textit{archetypes.csv} can join the simulation with an equal probability. If you wish to limit the number of types of agents that can walk around in the simulation and/or wish to change these probabilities, one can specify the `archetypes` and `weights` arguments of the constructor. For example, specifying:

```{r}
# Create instance of predped class with our office and two of the 
# default agent-types
my_predped <- predped(
  setting = my_background, 
  archetypes = c(
    "BaselineEuropean", 
    "DrunkAussie"
  ),
  weights = c(0.75, 0.25)
)
my_predped
```

states that you only want `"BaselineEuropean"`s and `"DrunkAussie"'s to walk around in the environment, where the former has a 75% probability of joining the environment while the latter only has a 25% probability of doing so. 



## Parameters 

Given the importance of knowing which agent-types exist by default, and given the importance of personalization in specifying the agent characteristics (see [_Advanced Simulations_](https://ndpvh.github.io/predped/articles/advanced_simulation.html)), it is useful to explain the parameter structure used by `predped`. 

To do so, we first load the default parameters of the package with the [`load_parameters`](https://ndpvh.github.io/predped/reference/load_parameters.html) function and inspect its type and structure:

```{r}
params <- load_parameters()

typeof(params)
names(params)
```

As one can see, the variable `params` is a named list containing the slots `"params_archetypes"`, `"params_sigma"`, and `"params_bounds"`. Each slot serves a different purpose in the specification of agents, as we detail below. 

First, the `"params_archetypes"` slot contains a `data.frame` connecting the names of the agent types with the parameters that belong to this type. The `data.frame` contains the following columns:

```{r}
colnames(params$params_archetypes)
```

The column `name` contains the name of the agent type, `color` contains the color that is used to visualize the agent in simulation plots, and `radius` determines the size of the agent. After these initial three columns follow parameters used by the utility functions on the operational level, namely:

- `randomness`: Controls the randomness of the decisions of the agent; 
- `stop_utility`: Controls the utility of stopping instead of moving, forming a threshold that moving options need to surpass;
- `reroute`: Controls the agent's tendency to reroute when their route is blocked by other agents;
- `a_turning`, `b_turning`: Controls the biomechanical limitations in the maintainence of speed while turning; 
- `preferred_speed`, `a_preferred_speed`, `b_preferred_speed`: Parameters that control the utility of continuing to move at the preferred speed; 
- `slowing_time`: Controls the time needed for the agent to slow down when approaching a goal; 
- `a_goal_direction`, `b_goal_direction`: Controls the extent to which agents will head directly in the direction of their goals; 
- `a_current_direction`, `b_current_direction`: Controls the extent to which agents will continue to head in their current direction; 
- `blr_current_direction`: Controls a left-right side bias when overtaking or crossing other agents;
- `a_interpersonal_distance`, `b_interpersonal_distance`, `d_interpersonal_distance`: Controls the extent to which agents leave room between them and other agents; 
- `a_blocked_angle`, `b_blocked_angle`: Controls the extent to which agents will anticipate and avoid directions that are blocked by other agents or objects at a distance; 
- `a_leader`, `b_leader`, `d_leader`: Controls the extent to which agents will follow a leader in crowded situations;
- `a_buddy`, `b_buddy`: Controls the extent to which agents will walk besides others of the ingroup:
- `a_group_centroid`, `b_group_centroid`: Controls the extent to which agents within a social group will tend to stick together;
- `b_visual_field`: Controls the extent to which agents within a social group will try to keep each other in a field of vision;
- `a_lgvf`, `b_lgvf`, `e_lgvf`: An improved social utility function that merges the principles of the group centroid and visual field utility functions described in the previous two bullet points.

The parameter names follow a particular convention, where for the typical utility function we use `a` is used for an exponent, `b` for a slope, and `d` for a difference between in-group and out-group members before the underscore `_`, and the name of the utility function after.

The values contained within this `data.frame` represent average values for the parameter for each agent type. For example, changing an agent type's `preferred_speed` to a higher value means that this type of agent will tend to walk faster. Yet, one typically expects agents within a particular type to show quantitative differences from their prototype. This variation is handled by the `"params_sigma"` slot in the `params` variable.

The `"params_sigma"` slot contains a named list of matrices, each slot of which is named after the different agent types and the contents of which define the (co)variation of each of the parameters. For example, the variation matrix for the `"BaselineEuropean"` looks like follows::

```{r}
head(params$params_sigma$BaselineEuropean)
```

Note that this matrix is <u>not</u> a covariance matrix, although for brevity will at times refer to it as such. Instead, it contains the standard deviations of each parameter on its diagonal and the correlations between each parameter on its off-diagonal, thus containing the following structure:

\begin{equation}
  \Gamma = \begin{bmatrix}
    \sigma_1 & \rho_{12} & \cdots & \rho_{1p} \\
    \rho_{21} & \sigma_{2} & \cdots & \rho{2p} \\
    \vdots & \vdots & \ddots & \vdots \\
    \rho_{p1} & \rho_{p2} & \hdots & \sigma_p \\
  \end{bmatrix}
\end{equation}

where $p$ is the number of parameters.

Computing this matrix can be achieved through the following equation, defining $\Lambda$ as a diagonal matrix containing the standard deviations of the parameters and $R$ as the matrix containing the correlations on its off-diagonal and $1$'s on its diagonal, we can construct the matrix $\Sigma$ as follows:

\begin{equation}
\begin{split}
    \Sigma &= \Lambda R \Lambda^T \\
    &= \Lambda R \Lambda
\end{split}
\end{equation}
where $T$ denotes the transpose.

We additionally note that the matrix $\Sigma$ does not act on the raw (bounded) values of the parameters defined by the `data.frame` in `"params_archetypes"`. Instead, it acts on a transformed unbounded version of these parameters, as explained below. The nature of this transformation falls outside of the scope of this vignette.

The final slot of the parameter list is `"params_bounds"`, a matrix containing the bounds of each of the parameters. By default, these bounds are the following:

```{r}
params$params_bounds
```

Lower and upper bounds for each of the parameters are contained in the first and second column respectively. Importantly, these bounds imply a fixed minimal and maximal value for each of the parameters, a feature that is preserved when generating random parameters for each agent. 

Because the process of generating parameters may be a little obscure, we allow users to visualize the variation in the parameters across agents through the [`plot_distribution`](https://ndpvh.github.io/predped/reference/plot_distribution.html) function, which typically will be used in the following way:

```{r}
#| fig.align = "center",
#| fig.dim = c(15, 15),
#| fig.alt = "One sees a series of histograms displaying the randomly drawn values of the parameters across 1000 iterations, using the mean, covariance, and bounds specified in the variables `means`, `covariances`, and `bounds`.",
#| warning = FALSE
# Save all separate components in separate variables
means <- params$params_archetypes
covariances <- params$params_sigma
bounds <- params$params_bounds

# Select the parameters of the BaselineEuropean
means <- means[means$name == "BaselineEuropean", ]
covariances <- covariances$BaselineEuropean

# Create a plot 
set.seed(1) # Set for reproducability purposes
plot_distribution(
  1000,
  mean = means,
  Sigma = covariances,
  bounds = bounds
)
```

For the default parameter settings, one can alternatively visualize the parameter distribution of a particular agent type as follows: 

```{r}
#| eval = FALSE
# Create a plot 
set.seed(1) # Set for reproducibility purposes
plot_distribution(
  1000,
  archetype = "BaselineEuropean"
)
```

which should output distributions similar to the previously plotted ones. Whichever way you use the function, it will output a set of histograms showing the variation in the values of each parameter with the current specifications of the three parameter slots. If you wish to play around with this, you can simply change the variables `means`, `covariance`, and `bounds` that were created earlier.



# Performing the Simulation

With the setup work done, we can finally go on to simulating pedestrian behavior as predicted by the M4MA. Having `my_predped` as the instance of the [`predped`](https://ndpvh.github.io/predped/reference/predped.html) class that combines our environment with several agent characteristics, we can perform a simulation through calling the [`simulate`](https://ndpvh.github.io/predped/reference/simulate.html) function:

```{r}
#| eval = TRUE
# Simulate behavior in our office for 100 iterations and maximally 3 agents
trace <- simulate(
  my_predped, 
  max_agents = 3, 
  iterations = 100
)
```

The variable `trace` is in fact an instance of the [`trace`](https://ndpvh.github.io/predped/reference/trace) class. This class contains all information relevant to the pedestrian behavior simulated by `predped`. It has the following attributes:

- `id`: An identifier for the trace;
- `time_step`: A numeric denoting the time step between each iteration;
- `setting`: The environment in which agents are walking around;
- `states`: An ordered list of lists containing the agents that are present in the environment at a particular iteration;
- `variables`: An ordered list of named lists containing variables that can be adjusted or used throughout the simulation (see [_Advanced Simulations_](https://ndpvh.github.io/predped/articles/advanced_simulation.html)).

Note that originally, the `trace` consisted of a list of instances of the [`state`](https://ndpvh.github.io/predped/reference/state) class, essentially containing all information about the agents, environment, and other variables that are relevant for the simulation at each iteration. Given that the underlying functions still make use of this class, it is useful to review its slots:

- `iteration`: The iteration number;
- `setting`: The environment in which the agents are walking around;
- `agents`: A list containing the agents that are currently present in the environment, each with their own characteristics, positions, orientations,...; 
- `potential_agents`: A list containing agents that are currently waiting at the entrance to enter the room;
- `variables`: A named list containing variables that can be adjusted or used during the simulation (see [_Advanced Simulations_](https://ndpvh.github.io/predped/articles/advanced_simulation.html));
- `iteration_variables`: A `data.frame` containing information for the simulation at each iteration. Is slowly being phased out.

To allow for backward compatibility, we provide functions that can transform one type of trace to the other, specifically [`trace_to_state`](https://ndpvh.github.io/predped/reference/trace_to_state) for transforming a `trace` to a list of `state`s and [`state_to_trace`](https://ndpvh.github.io/predped/reference/state_to_trace) for transforming a list of `state`s to a `trace`. Generally, you should not need to use these functions that often.

One can visualize both a single state of the simulation, or all states at once by calling the [`plot`](https://ndpvh.github.io/predped/articles/plot.html) function:

```{r}
#| fig.align = "center",
#| fig.alt = "The plot shows the environment "
# Plot a single iteration of the trace
plt <- plot(
  trace, 
  iterations = 10
)
plt
```

```{r}
#| eval = FALSE
# Plot the complete trace
plt <- plot(trace)
```

One can now skim through all plots to see how the agents behave, but this may turn out to be a tedious task. We therefore also recommend the creation of GIFs based on these plots. We recommend using the `gifski` package for this purpose:

```{r}
#| eval = FALSE
# Create a GIF that is 5 times faster than real time (10 frames per second 
# instead of 2)
gifski::save_gif(
  lapply(plt, print), 
  file.path("my_gif.gif"),
  delay = 1/10
)
```

In our experience, this is the quickest way of producing a GIF in R, but we warn the user: When a `trace` is long, this may still take a while. If you are experimenting with simulation settings before running a full-length simulation study, we recommend setting the `plot_live` argument of [`simulate`](https://ndpvh.github.io/predped/reference/simulate.html) to `TRUE` instead. Note that `plot_live` is experimental and is known to not function equally well across different IDE's. Setting a longer delay can sometimes help. 



# Recap

In summary, one can use the `predped` package to simulate data according to the M4MA by adhering to a particular workflow. Specifically, users need to specify (a) an environment, (b) the characteristics of the agents, and (c) the simulation characteristics. A minimal working example that combines all of these is shown the following code:

```{r}
#| eval = FALSE
# Load predped
library(predped)

# Create an environment: Simplified version of our office
my_background <- background(
  shape = rectangle(center = c(0, 0), size = c(4.5, 5)), 
  objects = list(
    # Desks
    rectangle(
      center = c(-0.85, 0), 
      size = c(2.4, 1.6),
      forbidden = 1
    ), 
    # Cabinets
    rectangle(
      center = c(-1, -2.3), 
      size = c(1.2, 0.4), 
      forbidden = c(1, 3, 4)
    ), 
    # Big bookcase
    rectangle(
      center = c(0.35, 2.275), 
      size = c(3, 0.45), 
      forbidden = c(1, 2, 3)
    )
  ), 
  entrance = c(2.25, 1.3)
)

# Link agent characteristics to the environment
my_predped <- predped(
  setting = my_background, 
  archetypes = c("BaselineEuropean", "DrunkAussie"),
  weights = c(0.5, 0.5)
)

# Simulate 120 iterations, or 1 minute
trace <- simulate(
  my_predped,
  max_agents = 3, 
  iterations = 120
)

# Plot the trace and save as a GIF
plt <- plot(trace)
gifski::save_gif(
  lapply(plt, print), 
  file.path("my_simulation.gif"),
  delay = 1/10
)
```
